---
title: "Class 17: Vaccination Rate Mini Project"
author: "Elena"
format: gfm
editor: visual
toc: TRUE
---

# Exploring Data

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```
## Q1. What column details the total number of people fully vaccinated?

persons_fully_vaccinated

## Q2. What column details the Zip code tabulation area?

zip_code_tabulation_area

## Q3. What is the earliest date in this dataset?

2021-01-05

## Q4. What is the latest date in this dataset?

2022-11-22 (seems that there are new entries since the latest one noted on 2022-11-15 for the online worksheet)

**Data Overview**

```{r}
skimr::skim(vax)
```

## Q5. How many numeric columns are in this dataset?

13

## Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

14921

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

##  Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

8.54%

```{r}
sum(is.na(vax$persons_fully_vaccinated)) / nrow(vax)
```

## Q8. [Optional]: Why might this data be missing?

Could be unreported or the data may be difficult to acquire.

**Dates**

```{r, message=F}
library(lubridate)
```

```{r}
today()
```

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
today() - vax$as_of_date[1]
```

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

## Q9. How many days have passed since the last update of the dataset?

5 days

```{r}
today() - max(vax$as_of_date)
```

## Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

99 (98 if including up to 2022-11-15 for the online worksheet)

```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP codes

```{r, message=F}
library(zipcodeR)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r}
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109"))
```

# Focus on SD

```{r}
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

## Q11. How many distinct zip codes are listed for San Diego County?

107

```{r}
length(unique(sd$zip_code_tabulation_area))
```

## Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

92154

```{r}
which.max(sd$age12_plus_population)
```

```{r}
sd$zip_code_tabulation_area[53]
```

## Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”?

0.74

```{r}
#Filter for 2022-11-15 data
vax.sd.1115 <- sd %>% filter(as_of_date == "2022-11-15")
```

```{r}
mean(vax.sd.1115$percent_of_population_fully_vaccinated, na.rm=T)
```

## Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
ggplot(vax.sd.1115) + 
  aes(percent_of_population_fully_vaccinated) + 
  geom_histogram(binwidth=0.1) +
  labs(title="Histogram of Vaccination Rates in San Diego County", x="Percent of Population Fully Vaccinated", y="Count")
```

# Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

## Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area

```{r}
ggplot(ucsd) +
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title="Vaccination rate for La Jolla CA 92037", x="Date", y="Percent Vaccinated")
```

## Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")
head(vax.36)
```

```{r}
mean.vax.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean.vax.36
```

```{r}
ggplot(ucsd) +
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title="Vaccination rate for La Jolla CA 92037", x="Date", y="Percent Vaccinated") +
  geom_hline(yintercept=mean.vax.36, col="red", lty="dashed")
```

## Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?

- Min 0.3785

- First Q 0.6396

- Median 0.7155

- Mean 0.7173

- Third Q 0.7880

- Max 1.0000

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

## Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  labs(x="Percent Vaccinated", y="Count") +
  xlim(0,1) + ylim(0,60)
```

## Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

Both 92109 (0.693299) and 92040 (0.546646) are below the average value (0.7172851)

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

## Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown.") +
  geom_hline(yintercept = mean.vax.36, linetype="dashed")
```

## Q21. How do you feel about traveling for Thanksgiving Break and meeting for in-person class afterwards?

Not so great! Seems like a lot of the vaccination rates in certain counties are below the average and have plateaued since July 2021. Folks could also travel to other parts of the country/world, where vaccination rates could vary. Finally, vaccination isn't (and shouldn't be) the sole protection against COVID-19 - other methods, such as wearing masks and ventilation, also play an important role in preventing the spread of the virus!

