---
title: "Class 11: Intro to Genome Informatics"
author: "Elena"
format: pdf
editor: visual
toc: TRUE
---

## Q5. What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)?

14.1%

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

## Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

- A/A: Sample Size 108, Median Expression 31.2

- A/G: Sample Size 233, Median Expression 25.1

- G/G: Sample Size 121, Median Expression 20.1

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r, message=F}
library(dplyr)
expr.aa <- expr %>% filter(geno=="A/A")
expr.ag <- expr %>% filter(geno=="A/G")
expr.gg <- expr %>% filter(geno=="G/G")
```

```{r}
median(expr.aa$exp)
```

```{r}
median(expr.ag$exp)
```

```{r}
median(expr.gg$exp)
```

## Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

G/G has a lower expression value relative to A/A, and A/G has an intermediate expression value; it seems that G reduces the expression of ORMDL3. From this observation, I would infer that the SNP does affect the expression of ORMDL3.

```{r, message=F}
library(ggplot2)
ggplot(expr) + 
  aes(x=geno,y=exp,fill=geno) + 
  geom_boxplot(show.legend=F, notch=T) +
  geom_jitter(show.legend=F, alpha=0.3) +
  labs(x="Genotype",y="Expression")
```
